\chapter{Sets as numbers -- some useful tricks}
Section*~\ref{sec:improving-performance} details a number of steps taken in order to build a performant Julia implementation of \pr{Knuth-Matroid}. Perhaps chief among these steps in terms of sheer performance gain compared to the initial, naïve implementation, was the transition from representing subsets of $E$ as a \jlinl{Set} of integers (or whatever type the elements of E might have), to representing them as a single integer, whose 1-bits denote which elements are in the set. This is possible as long as $n$ is less than the widest available integer type (in off-the-shelf Julia, 128 bits, though one can go wider with the help of libraries~\cite{BitIntegers.jl}). We reiterate the bitwise equivalents of the basic set operations in Table~\ref{tab:equiv-bitwise}.
\begin{table}[!ht]
  \caption{Set operations and their equivalent bitwise operations}
  \label{tab:equiv-bitwise}
  \centering
  \begin{tabular}{|l|l|}
  \hline
      Set operation   & Bitwise equivalent   \\\hline
      $A \cap B$      & $A \land B$       \\\hline
      $A \cup B$      & $A \lor B$        \\\hline
      $A \setminus B$ & $A \land \lnot B$   \\\hline
      $A \subseteq B$ & $A \land B$ = $A$ \\\hline
  \end{tabular}
\end{table}
These bitwise equivalents allow us to perform the set operations in constant time~(right???), resulting in significant performance increases. In the code snippets included throughout Section~\ref{sec:improving-performance} and Appendix~\ref{apx:code}, a number of ``tricks'' are performed with bitwise operations whose workings and purpose might be a bit obtuse. This appendix came to be as I got to grips with working with sets in this manner.

\section*{How do I...}
\subsection*{...create a singleton set?}
The left-shift operator ($<<$) can be used to set the $i$th bit to 1 and the others to 0. In general, $\{a\} = 1<<a$. This is used in an early version of \pr{Generate-Covers}:
\begin{jllisting}
function generate_covers_v2(F_r, n)
  Set([A | 1 << i for A ∈ F_r for i in 0:n-1 if A & 1 << i === 0])
end
\end{jllisting}

\subsection*{...find the smallest element of a set?}
Using the two's complement of a set $T$, denoted by $-T = \lnot T+1$, we can find the smallest element with the operation $T\land -T$. This is used in the next trick.

\subsection*{...enumerate all elements of a set one by one?}
Using the previous trick, we can repeatedly pop the smallest element in the following manner:
\begin{jllisting}
  t = 0b11111111
  while t > 0
    x = t&-t  # x is the singleton set consisting of the smallest element of t
    output(x)
    t &= ~x   # t = t setminus x
  end
\end{jllisting}
This outputs all numbers from 1 to 0xff with a Hamming weight of 1.

\subsection*{...get a random element from a set?}
We find all the positions at which the reversed bitstring of the set has a '1' character, and choose a random one.

\begin{jllisting}
function rand_el(S::Integer)
  x = rand([2^(i-1) for (i,c) in enumerate(reverse(bitstring(S))) if c == '1'])
  return convert(typeof(S), x)
end
\end{jllisting}